---
title: "Fitting a density surface model"
author: "Leonard Hambrecht, IRSS, University of British Columbia"
date: "September 2026"
bibliography: references.bib
---

## Introduction

A density surface model (DSM) is a two-stage model [@miller2013]. The [detection function](detection-function.qmd) corrects each segment's count for the animals that were missed. A generalized additive model (GAM) then relates the corrected counts to covariates, so that density can be predicted anywhere the covariates are known. In this article we fit DSMs for moose in Riding Mountain National Park, check them, and assess whether predicting over the whole park is an extrapolation.

We follow the habitat-based approach of @roberts2016. Density is modelled from habitat covariates only, without a spatial smooth of the coordinates, so the model describes *why* moose are where they are rather than just *where* they were seen. Each species gets its own model. For comparison we also fit a purely spatial model, as in the `dsm` example [*A simple DSM analysis: Gulf of Mexico dolphins*](https://distancesampling.org/dsm/articles/lines_gomex/mexico-analysis.html).

## Prerequisites

```{r}
#| label: packages
#| message: false
library(here)
library(dsm)      # density surface models (loads mgcv)
library(dsmextra) # extrapolation checks
library(sf)
library(ggplot2)
```

`dsmextra` is not on CRAN; install it from GitHub with `remotes::install_github("hambrecht/dsmextra")`. These articles were built with a [fork of `dsm`](https://github.com/hambrecht/dsm) that includes speed improvements; the CRAN version gives the same results (see [Packages](index.qmd#packages)).

We load the segment, observation and prediction tables from [*Adding covariates and building a prediction grid*](covariates.qmd), and the detection function from [*Estimating the detection function*](detection-function.qmd).

```{r}
#| label: load
model_data <- readRDS(here("workflow", "outputs", "model_data.rds"))
list2env(model_data, envir = environment())
df_best <- readRDS(here("workflow", "outputs", "detection_function.rds"))
```

The observation table holds all species. For the moose model we keep only moose sightings, while the detection function, fitted to all species, still provides each sighting's detection probability.

```{r}
#| label: moose
obs_moose <- obs[obs$Species == "Moose", ]
nrow(obs_moose)
mean(!segs$Sample.Label %in% obs_moose$Sample.Label)
```

Almost 98% of segments have no moose. This is typical of aerial surveys and it shapes the choice of response distribution.

## Response and response distribution

Because the detection function includes covariates, each sighting has its own detection probability. The response is therefore `abundance.est`: the count in each segment divided by the detection probabilities of its sightings (a Horvitz–Thompson estimate). This is non-negative and continuous, with a large mass of exact zeros.

The Tweedie distribution (`tw()`) models exactly this: it has a point mass at zero and a continuous density above zero, with its power parameter estimated from the data. The alternative, quasi-Poisson, is a mean–variance relationship for integer counts rather than a full distribution, and it has no likelihood, so it cannot be compared by AIC. Tweedie was also the choice of @roberts2016 for zero-inflated line-transect data.

All models are fitted by REML with `gamma = 1.4`, which penalises wiggliness a little more than the default and guards against overfitting [@wood2017].

## A spatial model

The simplest DSM smooths abundance over the coordinates of the segment centroids. It shows where moose were seen but explains nothing about why.

```{r}
#| label: dsm-xy
dsm_xy <- dsm(abundance.est ~ s(x, y, bs = "ts"), df_best, segs, obs_moose,
              family = tw(), gamma = 1.4, method = "REML")
summary(dsm_xy)
```

```{r}
#| label: fig-dsm-xy
#| fig-cap: "Fitted spatial smooth for moose (link scale)."
vis.gam(dsm_xy, view = c("x", "y"), plot.type = "contour", too.far = 0.05,
        main = "", asp = 1)
```

## Habitat models

We have two candidate habitat covariates: canopy cover (`pabove2`) and canopy height (`zmax`). Each enters as a thin-plate shrinkage smooth (`bs = "ts"`) with `k = 5`, which caps each smooth at about four degrees of freedom. This keeps the fitted responses simple enough to interpret and to trust outside the exact conditions sampled. The shrinkage basis can shrink a smooth all the way to zero, so an unhelpful covariate effectively drops out.

```{r}
#| label: dsm-habitat
dsm_p2 <- dsm(abundance.est ~ s(pabove2, bs = "ts", k = 5), df_best, segs, obs_moose,
              family = tw(), gamma = 1.4, method = "REML")
dsm_zmax <- dsm(abundance.est ~ s(zmax, bs = "ts", k = 5), df_best, segs, obs_moose,
                family = tw(), gamma = 1.4, method = "REML")
dsm_both <- dsm(abundance.est ~ s(pabove2, bs = "ts", k = 5) + s(zmax, bs = "ts", k = 5),
                df_best, segs, obs_moose,
                family = tw(), gamma = 1.4, method = "REML")
```

### Concurvity

Canopy cover and canopy height both describe the structure of the forest, so they are related. In a GAM this is called concurvity, the non-linear version of collinearity. When two smooths are concurve, their individual shapes become unstable and their uncertainty is inflated, even if the combined prediction looks reasonable. Values close to 1 are a warning sign.

```{r}
#| label: concurvity
cor(segs$pabove2, segs$zmax)
concurvity(dsm_both, full = FALSE)$estimate
```

Here the concurvity is moderate (about 0.5). With the full set of lidar metrics, however, cover and height metrics reached concurvity of 0.98 in the RMNP analysis, so every cover metric was paired with a height metric in a banned list, and only one of the two families may enter a model. We keep that rule here. Following @roberts2016, we choose the covariate that explains more deviance on its own, and then retain it only if its smooth is significant ($p \le 0.05$) and not shrunk away (effective degrees of freedom, EDF, of at least 0.85).

```{r}
#| label: compare-habitat
data.frame(
  model = c("s(pabove2)", "s(zmax)"),
  dev_expl = c(summary(dsm_p2)$dev.expl, summary(dsm_zmax)$dev.expl),
  AIC = c(AIC(dsm_p2), AIC(dsm_zmax))
)
```

Canopy height explains more than twice as much deviance as canopy cover. Neither explains much: a few percent of deviance explained is common for DSMs of sparse sightings, where most of the variation between segments is the chance of seeing an animal at all.

```{r}
#| label: select-habitat
dsm_hab <- if (summary(dsm_p2)$dev.expl >= summary(dsm_zmax)$dev.expl) dsm_p2 else dsm_zmax
summary(dsm_hab)
```

```{r}
#| label: fig-smooth
#| fig-cap: "Estimated effect of the selected habitat covariate on moose abundance (link scale), with the covariate values of the segments as a rug."
plot(dsm_hab, shade = TRUE, rug = TRUE, residuals = FALSE, pages = 1)
```

## Model checking

`gam.check()` produces four standard diagnostic plots and a check of the basis size $k$. For count data with many zeros, the Q–Q plot and the residuals against the linear predictor show striping from the zeros, which makes them hard to read.

```{r}
#| label: fig-gam-check
#| fig-cap: "Standard GAM diagnostics for the moose habitat model."
#| fig-width: 8
#| fig-height: 8
gam.check(dsm_hab)
```

Randomized quantile residuals avoid that problem [@dunn1996]. If the model is adequate they should scatter evenly around zero, with no pattern against the linear predictor.

```{r}
#| label: fig-rqgam
#| fig-cap: "Randomized quantile residuals against the linear predictor."
rqgam_check(dsm_hab)
```

## Extrapolation

A DSM is only reliable where the prediction grid resembles the conditions that were surveyed. `dsmextra` identifies prediction cells whose covariate values lie outside the range of the segments (univariate extrapolation) or in combinations not seen in the segments (combinatorial extrapolation) [@bouchet2020].

```{r}
#| label: extrapolation
#| message: false
covs <- c("pabove2", "zmax")
extrap <- compute_extrapolation(
  samples = segs[, covs],
  covariate.names = covs,
  prediction.grid = st_drop_geometry(pred)[, c("x", "y", covs)],
  coordinate.system = sp::CRS(SRS_string = paste0("EPSG:", LOCAL))
)
summary(extrap)
```

```{r}
#| label: fig-extrapolation
#| fig-cap: "Extrapolation (ExDet) across the prediction grid. Values below 0 are univariate extrapolation, values above 1 combinatorial; values between 0 and 1 are within the sampled conditions."
pred$ExDet <- extrap$data$all$ExDet
ggplot(pred) +
  geom_sf(aes(fill = ExDet), colour = NA) +
  scale_fill_gradient2(low = "#b2182b", mid = "grey95", high = "#2166ac", midpoint = 0.5) +
  theme_minimal()
```

The flight lines covered the park systematically, so almost all prediction cells fall within the range of sampled canopy conditions.

## Other species

The same procedure applies to the other species. Following @roberts2016, the model's complexity depends on how often a species was seen. Species with at least 40 sightings get a habitat model chosen as above. Species with fewer than 20 get an intercept-only model: this estimates the average density across the study area without trying to map it.

```{r}
#| label: fit-species
fit_species <- function(species, covariates = c("pabove2", "zmax")) {
  obs_sp <- obs[obs$Species == species, ]
  fit <- function(rhs) {
    dsm(as.formula(paste("abundance.est ~", rhs)), df_best, segs, obs_sp,
        family = tw(), gamma = 1.4, method = "REML")
  }
  if (nrow(obs_sp) < 20) return(fit("1"))

  # One covariate (they are concurve): the one with most deviance explained
  singles <- lapply(covariates, function(v) fit(sprintf("s(%s, bs = 'ts', k = 5)", v)))
  best <- singles[[which.max(sapply(singles, function(m) summary(m)$dev.expl))]]

  # Keep it only if significant and not shrunk away
  st <- summary(best)$s.table
  if (st[1, "p-value"] <= 0.05 && st[1, "edf"] >= 0.85) best else fit("1")
}

species_list <- c("Moose", "Deer", "Elk", "Bison", "Wolf")
dsm_species <- lapply(setNames(species_list, species_list), fit_species)
sapply(dsm_species, function(m) deparse(formula(m)))
```

Canopy height is retained for moose and deer. For elk neither canopy metric meets the selection criteria, so elk also get an intercept-only model. Covariates other than the two used here would likely help. In the full RMNP analysis, which also considered elevation and distance to forest edge, elevation was the strongest predictor for all three species.

## Saving

```{r}
#| label: save
saveRDS(list(dsm_species = dsm_species, dsm_xy = dsm_xy),
        here("workflow", "outputs", "dsm_models.rds"))
```

**Next:** [Estimating abundance and its uncertainty](abundance.qmd).

## References

::: {#refs}
:::
