---
title: "Survey designs with subplots"
author: "Leonard Hambrecht, IRSS, University of British Columbia"
date: "September 2026"
bibliography: references.bib
---

## Introduction

A crewed aircraft can fly transects across a whole study area in one or a few long flights. A drone is limited by its endurance and by how far it may fly from its operators. One practical design is therefore to divide the effort into *subplots*: compact blocks, each small enough to survey in one flight, with transects placed inside them. The Alberta simulation study compared such designs for fixed-wing drones and quadcopters with the conventional segmented grid design, matching the total length of transect flown ([*Multi-platform simulations*](../sim_multiplatform.qmd)).

The `dsims` examples cover designs across whole regions and designs with strata that divide a region completely ([*Grouped strata*](https://distancesampling.org/dsims/articles/dsims_grouped_strata.html)). This article shows how to generate subplots automatically, turn them into a region with one stratum per subplot, place systematic and zigzag transects inside them, and restrict the simulated population to the subplots. We compare them with two designs across the whole park: the conventional segmented grid of 10 km transects, which the following articles use as their default, and systematic lines with the same effort as the 2025 survey.

## Prerequisites

```{r}
#| label: packages
#| message: false
library(here)
library(dsims)
library(sf)
library(dplyr)
library(ggplot2)
```

Generating transects and coverage grids is much faster with the [fork of `dssd`](https://github.com/hambrecht/dssd) (see [Packages](index.qmd#packages)); the results are the same with the CRAN version.

We load the region, density grid and population from [*From density surface model to simulation*](simulation-density.qmd).

```{r}
#| label: load
setup <- readRDS(here("workflow", "outputs", "simulation_setup.rds"))
list2env(setup, envir = environment())
```

## Coverage grid

To compare designs we need to know how evenly each one covers the study area. `run.coverage()` generates many sets of transects and records, for each point of a coverage grid, the proportion of sets in which it fell inside the searched strip. A 2 km grid is fine enough for this.

```{r}
#| label: coverage-grid
cover <- make.coverage(region, spacing = 2000)
```

## A segmented grid design

The standard design for aerial ungulate surveys in Alberta is a sample of 10 km transects, preferably oriented north–south, that deals appropriately with edge effects [@aep2016]. The protocol describes a simple random sample of transects, and a systematic layout for small or geometrically complex study areas. `dssd` provides the systematic version as a *segmented grid*: 10 km segments on north–south lines 1.2 km apart, with a random start. `seg.threshold = 10` discards segments that are less than 10% (1 km) inside the park, and `edge.protocol = "minus"` handles the edge.

```{r}
#| label: design-seg
design_seg <- make.design(
  region = region, transect.type = "line", design = "segmentedgrid",
  spacing = 1200, seg.length = 10000, seg.threshold = 10, design.angle = 0,
  edge.protocol = "minus", truncation = TRUNCATION_M, coverage.grid = cover
)
transects_seg <- generate.transects(design_seg)
```

```{r}
#| label: fig-design-seg
#| fig-cap: "One set of transects for the segmented grid design."
plot(region, transects_seg)
```

This is the default design in the [following articles](simulation-detectability.qmd). Its effort is set by the spacing and segment length rather than matched to the 2025 survey, so it flies less transect than the design below.

## A design across the whole park

The 2025 survey flew `r round(effort / 1000)` km on effort over the park. As a baseline, we use systematic north–south lines across the whole park with the spacing that gives the same effort: the park area divided by the line length.

```{r}
#| label: design-full
spacing_full <- round(region@area / effort, -1)
spacing_full

design_full <- make.design(
  region = region, transect.type = "line", design = "systematic",
  spacing = spacing_full, design.angle = 0, edge.protocol = "minus",
  truncation = TRUNCATION_M, coverage.grid = cover
)
transects_full <- generate.transects(design_full)
```

```{r}
#| label: fig-design-full
#| fig-cap: "One set of transects for the design across the whole park."
plot(region, transects_full)
```

## Designing subplots

### Number and size of subplots

Each subplot is one flight. The Superwake SW-117 flies about 367 km in a 6-hour flight at 17 m/s, so the number of subplots is the total effort divided by one flight's length. Within a subplot the lines are spaced twice the truncation distance apart, so the searched strips of neighbouring lines meet without gaps or overlap.

```{r}
#| label: subplot-numbers
FLIGHT_LENGTH_M <- 17 * 6 * 3600 # 367.2 km per flight
n_plots <- round(effort / FLIGHT_LENGTH_M)
spacing_sub <- 2 * TRUNCATION_M
c(n_plots = n_plots, spacing = spacing_sub)
```

`find_best_block_dim()` finds the subplot shape closest to a square that fits a whole number of lines at this spacing within each subplot's share of the effort.

```{r}
#| label: block-dim
find_best_block_dim <- function(total_length, number_blocks, spacing) {
  block_length <- total_length / number_blocks
  possible_lines <- seq(1, block_length / spacing, 1)
  tibble(
    nr_lines = possible_lines,
    y_length = block_length / possible_lines,
    x_length = spacing * possible_lines,
    difference = abs(block_length / possible_lines - spacing * possible_lines)
  ) |>
    filter(total_length >= y_length * nr_lines * number_blocks) |>
    arrange(difference) |>
    slice(1)
}

plot_dim <- find_best_block_dim(effort, n_plots, spacing_sub)
plot_dim
```

### Placing the subplots

To spread the subplots evenly, we use `make.coverage()` in a different role: asked for a given number of grid points, it places them evenly over a region. We place the centres in the park shrunk by 1 km, so no subplot is centred right at the boundary.

```{r}
#| label: centres
inner <- make.region(region.name = "inner", shape = st_buffer(region@region, -1000), units = "m")
centres <- make.coverage(inner, n.grid.points = n_plots)@grid
nrow(centres)
```

`make.coverage()` can return one or two more points than requested. We build a rectangle around each centre, clip it to the park, and keep the `n_plots` largest.

```{r}
#| label: subplots
make_subplots <- function(centres, width, height) {
  xy <- st_coordinates(centres)
  polys <- lapply(seq_len(nrow(xy)), function(i) {
    st_polygon(list(cbind(
      xy[i, 1] + c(-1, 1, 1, -1, -1) * width / 2,
      xy[i, 2] + c(-1, -1, 1, 1, -1) * height / 2
    )))
  })
  st_sf(geometry = st_sfc(polys, crs = st_crs(centres)))
}

plots <- make_subplots(centres, plot_dim$x_length, plot_dim$y_length) |>
  st_intersection(st_geometry(region@region))
plots <- plots[order(st_area(plots), decreasing = TRUE)[seq_len(n_plots)], ]
plots$ID <- sprintf("plot%02d", seq_len(nrow(plots)))
```

### Subplots as a region

In `dsims`, each subplot becomes a stratum of a new region. Transects are then generated separately within each stratum.

```{r}
#| label: subplot-region
sub_region <- make.region(region.name = "subplots", shape = plots,
                          strata.name = plots$ID, units = "m")
sum(sub_region@area) / region@area
```

The subplots cover about `r round(100 * sum(sub_region@area) / region@area)`% of the park.

## Transects within subplots

We compare two designs within the subplots. Systematic parallel lines fill each subplot as planned. An equal-spaced zigzag (`eszigzag`) reduces the flying time between lines, because the aircraft does not have to turn back at the end of each line. We give the zigzag the same line length per subplot as the systematic design, taken from one set of systematic transects.

```{r}
#| label: design-subplots
design_sys <- make.design(
  region = sub_region, transect.type = "line", design = "systematic",
  spacing = spacing_sub, design.angle = 0, edge.protocol = "minus",
  truncation = TRUNCATION_M, coverage.grid = cover
)
transects_sys <- generate.transects(design_sys)

design_zz <- make.design(
  region = sub_region, transect.type = "line", design = "eszigzag",
  line.length = transects_sys@line.length, design.angle = 0, edge.protocol = "minus",
  truncation = TRUNCATION_M, coverage.grid = cover
)
transects_zz <- generate.transects(design_zz)
```

```{r}
#| label: fig-design-subplots
#| fig-cap: "Systematic (left) and zigzag (right) transects within the subplots."
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2))
plot(region, transects_sys)
plot(region, transects_zz)
```

## Coverage and effort

`run.coverage()` repeats the transect generation and summarises coverage, line length and total flight distance across repetitions. We use 100 repetitions per design, run in parallel (`run.parallel` is available in the `dssd` fork).

```{r}
#| label: run-coverage
#| cache: true
#| message: false
set.seed(2025)
design_full <- run.coverage(design_full, reps = 100, run.parallel = TRUE)
design_sys <- run.coverage(design_sys, reps = 100, run.parallel = TRUE)
design_zz <- run.coverage(design_zz, reps = 100, run.parallel = TRUE)
design_seg <- run.coverage(design_seg, reps = 100, run.parallel = TRUE)
```

The design statistics distinguish three distances: `line.length` is the length of transect on effort; `trackline` adds the travel between transects; and `cyclictrackline` adds the return from the last transect to the start. Coverage (`p.cov.area`) is the percentage of the region inside the searched strips.

```{r}
#| label: design-metrics
design_metrics <- function(design, name) {
  s <- design@design.statistics
  data.frame(
    design = name,
    on_effort_km = s$line.length["Mean", "Total"] / 1000,
    trackline_km = s$trackline["Mean", "Total"] / 1000,
    cyclic_km = s$cyclictrackline["Mean", "Total"] / 1000,
    on_effort_pct = 100 * s$line.length["Mean", "Total"] / s$cyclictrackline["Mean", "Total"],
    covered_pct = s$p.cov.area["Mean", "Total"]
  )
}

metrics <- rbind(
  design_metrics(design_full, "whole park, systematic"),
  design_metrics(design_sys, "subplots, systematic"),
  design_metrics(design_zz, "subplots, zigzag"),
  design_metrics(design_seg, "segmented grid")
)
metrics$covered_pct[2:3] <- metrics$covered_pct[2:3] * sum(sub_region@area) / region@area
knitr::kable(metrics, digits = 1)
```

Coverage for the subplot designs is given as a percentage of the park rather than of the subplots, so all rows are comparable. Four points stand out:

- **Less effort than planned.** The subplot designs fly about `r round(100 * (1 - 1000 * metrics$on_effort_km[2] / effort))`% less transect than planned, because the park boundary clips some subplots. This was a common difficulty in the Alberta study as well: fixed-size subplots rarely fit an irregular study area exactly.
- **Less of the park covered.** With less effort and all of it inside the subplots, the subplot designs cover `r round(metrics$covered_pct[2])`% of the park, against `r round(metrics$covered_pct[1])`% for the whole-park design.
- **More flying on effort with the zigzag.** The zigzag spends a larger share of the flight on effort than systematic lines in the same subplots (`r round(metrics$on_effort_pct[3])`% against `r round(metrics$on_effort_pct[2])`%), because it avoids the turns between parallel lines.
- **The segmented grid.** The segmented grid flies `r round(metrics$on_effort_km[4])` km on effort and covers `r round(metrics$covered_pct[4])`% of the park. Because its segments are short, `r round(100 - metrics$on_effort_pct[4])`% of its flight is spent travelling between them.

```{r}
#| label: fig-coverage
#| fig-cap: "Coverage scores for the whole-park design: the proportion of repetitions in which each grid point was inside a searched strip."
plot(design_full)
```

## Restricting the population to the subplots

`dsims` estimates density for the region it is given. A design confined to subplots cannot provide an estimate for the whole park, so the population in the simulation must also be confined to the subplots: the estimates are then compared with the true density *inside the subplots*. We therefore cut the moose density grid to the subplots, label each piece with its subplot as the stratum, and set the number of groups in each stratum to its expected abundance under the DSM divided by the mean group size.

```{r}
#| label: density-subplots
density_moose <- densities$Moose
cells_sub <- st_intersection(density_moose@density.surface[[1]], plots) |>
  mutate(strata = ID) |>
  select(strata, density, x, y)

density_sub <- density_moose
density_sub@density.surface <- list(cells_sub)
density_sub@region.name <- sub_region@region.name
density_sub@strata.name <- sub_region@strata.name

N_sub <- tapply(cells_sub$density * as.numeric(st_area(cells_sub)), cells_sub$strata, sum)
N_sub <- round(N_sub[sub_region@strata.name] / group_size[["Moose"]])
N_sub
```

`N_sub` is ordered to match the strata of the region, because `make.population.description()` matches the population sizes to the strata by position.

```{r}
#| label: population-subplots
pop_desc_sub <- make.population.description(
  region = sub_region, density = density_sub,
  covariates = list(size = list(distribution = "ztruncpois", mean = group_size[["Moose"]])),
  N = as.numeric(N_sub), fixed.N = TRUE
)
```

## Simulating the designs

Now we can run each design many times and see how well it estimates moose density. For this comparison, detectability is constant: the moose detection function at the median canopy cover (see [*From density surface model to simulation*](simulation-density.qmd#defining-detectability)). The next article makes detectability depend on canopy.

Each simulated survey is analysed with a hazard-rate detection function without covariates. For the subplot designs, the subplots are strata of the design but are analysed together, like the grouped strata in the `dsims` article [*Grouped strata*](https://distancesampling.org/dsims/articles/dsims_grouped_strata.html). `group.strata` maps every subplot to a single analysis stratum.

```{r}
#| label: analyses
analysis <- make.ds.analysis(dfmodel = ~1, key = "hr", truncation = TRUNCATION_M)
analysis_sub <- make.ds.analysis(
  dfmodel = ~1, key = "hr", truncation = TRUNCATION_M,
  group.strata = data.frame(design.id = sub_region@strata.name, analysis.id = "subplots")
)
```

Each of the 100 repetitions generates a new population, a new set of transects and new detections, and fits the detection function. The simulations run in parallel and take a few minutes each; for your own designs, start with 20 repetitions to check the setup.

```{r}
#| label: simulate-designs
#| cache: true
#| warning: false
REPS <- 100
set.seed(2025)
run <- function(design, pop, analysis) {
  sim <- make.simulation(reps = REPS, design = design, population.description = pop,
                         detectability = detect_constant, ds.analysis = analysis)
  run.simulation(sim, run.parallel = TRUE)
}
sims <- list(
  "segmented grid" = run(design_seg, pop_descs$Moose, analysis),
  "whole park, systematic" = run(design_full, pop_descs$Moose, analysis),
  "subplots, systematic" = run(design_sys, pop_desc_sub, analysis_sub),
  "subplots, zigzag" = run(design_zz, pop_desc_sub, analysis_sub)
)
```

We compare the designs on density, because the subplot designs estimate density within the subplots, not across the whole park. For each design we extract the true and mean estimated density of individuals (animals per km²), the bias, the root mean squared error relative to the true density (RRMSE), the proportion of 95% confidence intervals that contain the true density, and the mean coefficient of variation.

```{r}
#| label: design-results
sim_metrics <- function(sim) {
  D <- summary(sim, description.summary = FALSE)@individuals$D
  D <- D[nrow(D), ] # the total, or the single analysis stratum
  data.frame(true_D = D$Truth * 1e6, mean_D = D$mean.Estimate * 1e6,
             bias_pct = D$percent.bias, rrmse = D$RMSE / D$Truth,
             ci_coverage = D$CI.coverage.prob, mean_cv = D$mean.se / D$mean.Estimate)
}
design_results <- cbind(design = names(sims), do.call(rbind, lapply(sims, sim_metrics)))
rownames(design_results) <- NULL
knitr::kable(design_results, digits = 3)
```

With 100 repetitions, the Monte Carlo standard error of the bias is about one percentage point, so only larger differences are meaningful. The bias ranges from `r round(min(design_results$bias_pct), 1)`% to `r round(max(design_results$bias_pct), 1)`% and the RRMSE from `r round(min(design_results$rrmse), 2)` to `r round(max(design_results$rrmse), 2)`. Precision follows effort. The whole-park systematic design flies the most transect and has the lowest RRMSE (`r round(design_results$rrmse[design_results$design == "whole park, systematic"], 2)`). The segmented grid flies the least and has the highest RRMSE (`r round(design_results$rrmse[design_results$design == "segmented grid"], 2)`). The subplot designs lie in between.

Two caveats apply:

- **The estimate is for the subplots, not the park.** To estimate abundance for the whole park from subplots, the survey data have to be modelled with a DSM, predicting from habitat covariates as in [*Fitting a density surface model*](dsm.qmd).
- **The subplots here happen to be representative.** Their true density is close to the park's. Subplots placed in unrepresentative habitat would estimate a different density.

## Saving

```{r}
#| label: save
saveRDS(
  list(design_seg = design_seg, design_full = design_full, design_sys = design_sys, design_zz = design_zz,
       sub_region = sub_region, pop_desc_sub = pop_desc_sub, plots = plots,
       metrics = metrics, design_results = design_results),
  here("workflow", "outputs", "simulation_designs.rds")
)
```

**Next:** [Simulating spatially varying detectability](simulation-detectability.qmd).

## References

::: {#refs}
:::
