---
title: "Interactive Map Example"
format: html
page-layout: full
execute: 
  echo: false
  warning: false
  message: false
---
| [Home](index.html) | [Simulations](sim.html) | [Deep Learning](dl.html) | [Field Work](fieldwork.html) | [IRSS](https://irsslab.forestry.ubc.ca/){target="_blank"} | [fRI Research](https://friresearch.ca/){target="_blank"} | [Contact](mailto:137965368+hambrecht@users.noreply.github.com) |
# Thunderbay
The aim of this field campaign is to collect thermal images of moose using a fixed-wing drone under operationally realistic conditions. The data will be analyzed to establish the detection function and evaluate how different environmental variables influence detection probability. In addition, the collected images will be used to train a deep-learning model for moose detection and identification.
The study area, located north of Thunder Bay, Ontario, was chosen because the Ontario Ministry of Natural Resources (MNR) has over 100 GPS-collared moose, providing reliable ground-truth data. The western portion of the study area includes active logging sites with various treatment options, as well as protected areas such as a provincial park and a conservation reserve. This landscape offers a wide range of canopy and snow cover conditions. The terrain features only minor elevation changes; however, Dog Lake, a prominent feature at the centre of the study area, is frozen during winter. Between the lake and the more sheltered forest stands, we expect significant temperature variation. Logging roads provide convenient access to most parts of the area.
The combination of GPS-collared moose, diverse environmental conditions, and good accessibility makes the study area particularly well suited for this project. The data collected will be used to assess how environmental factors—especially canopy cover, but also snow cover and temperature—affect the probability of detecting moose in drone-based thermal imagery. Importantly, the precise location data from GPS collars provided by MNR make this dataset unique and enable a more robust assessment of detection probability.

## Flight plan specifications
Flight Details:
- Transect lines oriented North–South, spaced 600 m apart.
- Flight altitude: 120 m
- Speed: 65 km/h
- Priorities thermal photo for deep-learning
  - Optimized gimbal movements to reduce motion blur in images
  - No on-screen display (OSD) in images
- Only zoom in far enough to confirm species with RGB camera
- 
A two-week window (10 working days) is expected to provide sufficient opportunity for data collection. Two members from the University of British Columbia (UBC) will be present in the field to coordinate operations.

## My Interactive Map

```{r include=FALSE}
library(here)
library(leaflet)
library(leafem)
library(sf)
library(terra)
library(dplyr)
library(RColorBrewer)
privateLandImage <- rast("mapLayers/images/image003_WGS84.tif")
fracSnow <- rast("mapLayers/LC08_CU_020002_20250311_20250328_02_VIEWABLE_SNOW_MASKED_SCALED_SMALLEST.TIF")
LAI2021 <- rast("mapLayers/LAI/custom_region_LAI_2021Winter_mean_scaled_uint16.tif")
LAI2122 <- rast("mapLayers/LAI/custom_region_LAI_2122Winter_mean_scaled_uint16.tif")
LAI2223 <- rast("mapLayers/LAI/custom_region_LAI_2223Winter_mean_scaled_uint16.tif")
LAI2324 <- rast("mapLayers/LAI/custom_region_LAI_2324Winter_mean_scaled_uint16.tif")
LAI2425 <- rast("mapLayers/LAI/custom_region_LAI_2425Winter_mean_scaled_uint16.tif")
forestLoss <- rast("mapLayers/Hansen_GFC-2024-v1.12_lossyear_50N_090WMASKED.tif")
forestGain <- rast("mapLayers/Hansen_GFC-2024-v1.12_gain_50N_090WMASKED.tif")
studyArea <- st_read("mapLayers/studyAreaOntario.shp")
studyArea <- st_transform(studyArea, crs = 4326) # WGS84
surveyArea <- st_read("mapLayers/surveyArea.shp")
surveyArea <- st_transform(surveyArea, crs = 4326) # WGS84
aerodrome <- st_read("mapLayers/aerodrome.shp")
aerodrome <- st_transform(aerodrome, crs = 4326) # WGS84
canopyCover <-st_read("mapLayers/canopyCover.shp")
canopyCover <- st_transform(canopyCover, crs = 4326) # WGS84
canopyCover$STRUCTUR_9 <- round(canopyCover$STRUCTUR_9,0)
buildings <-st_read("mapLayers/OntarioGeoHub/buildingsClipped.shp")
buildings <- st_transform(buildings, crs = 4326) # WGS84
township <-st_read("mapLayers/OntarioGeoHub/GeographicTownshipClipped.shp")
township <- st_transform(township, crs = 4326) # WGS84
mooseWinterAreas <-st_read("mapLayers/OntarioGeoHub/mooseWinterAreas.shp")
mooseWinterAreas <- st_transform(mooseWinterAreas, crs = 4326) # WGS84
provincalPark <-st_read("mapLayers/OntarioGeoHub/provinalParkClipped.shp")
provincalPark <- st_transform(provincalPark, crs = 4326) # WGS84
recreationPoints <-st_read("mapLayers/OntarioGeoHub/recreationPointsClipped.shp")
recreationPoints <- st_transform(recreationPoints, crs = 4326) # WGS84
towers <-st_read("mapLayers/OntarioGeoHub/towersClipped.shp")
towers <- st_transform(towers, crs = 4326) # WGS84
lake <-st_read("mapLayers/OSM/dogLake.shp")
lake <- st_transform(lake, crs = 4326) # WGS84
roads <-st_read("mapLayers/OSM/road.shp")
roads <- st_transform(roads, crs = 4326) # WGS84


# Define a color palette based on the 'name' column

palElevation <- colorNumeric(
  palette = brewer.pal(9, "YlGnBu"),
  domain = c(0, 600)  # Set min and max values
)
palGreen <- colorNumeric(
  palette = brewer.pal(n = 9, name = "YlGn"),
  domain = canopyCover$STRUCTUR_9
)
palSnow <- colorNumeric(
  palette = rev(brewer.pal(n = 9, name = "PuBu")),
  domain =  range(values(fracSnow), na.rm = TRUE),
  na.color = "transparent"   # make NAs transparent
)
palLAI <- colorNumeric(
  palette = brewer.pal(n = 9, name = "YlGnBu"),
  domain = c(0,46500),
  na.color = "transparent"   # make NAs transparent
)
palLoss <- colorNumeric(
  palette = brewer.pal(n = 9, name = "Reds"),
  domain = range(values(forestLoss), na.rm = TRUE),
  na.color = "transparent"
)
palGain <- colorNumeric(
  palette = brewer.pal(n = 2, name = "Greens"),
  domain = range(values(forestGain), na.rm = TRUE),
  na.color = "transparent"
)
```
```{r, include=FALSE}
range(values(forestGain), na.rm = TRUE)

```
```{r leaflet-map, echo=FALSE}
#| echo: false
#| out-width: "100%"
#| out-height: "60vh"
#| column: page
leaflet() %>%
#Base layers
  addTiles() %>%
  # missing layer: https://www.gisapplication.lrc.gov.on.ca/CLUPA/Index.html?site=CLUPA&viewer=CLUPA&locale=en-US
  addTiles(
    "https://tile.opentopomap.org/{z}/{x}/{y}.png",
    layerId = "_14b46c75_d6e7_481f_b57d_a0bbb927294a",
    group = "Base layer: OpenTopo",
    options = WMSTileOptions()
  ) %>%
  addTiles(
    "https://s3.amazonaws.com/elevation-tiles-prod/terrarium/{z}/{x}/{y}.png",
    layerId = "_e5657323_f251_4d70_9366_f5879b155d3e",
    group = "Base layer: Mapzen Global Terrain",
    options = WMSTileOptions()
  ) %>%
  addTiles(
    "https://datacube.services.geo.ca/wrapper/ogc/elevation-hrdem-mosaic",
    layerId = "Relief_map_of_the_Digital_Terrain_Model____________________Carte_de_relief_du_Mod_le_Num_rique_de_Terrain__139f2d65_7f2c_44b9_834a_01214bd59908",
    group = "Base layer: Relief map of DTM",
    options = WMSTileOptions(),
    attribution = "© Natural Resources Canada"
  ) %>%
  addTiles(
    urlTemplate = "https://ws.lioservices.lrc.gov.on.ca/arcgis2/rest/services/LIO_Imagery/Ontario_Imagery_Web_Map_Service/MapServer/WMTS/tile/1.0.0/LIO_Imagery_Ontario_Imagery_Web_Map_Service/default/GoogleMapsCompatible/{z}/{y}/{x}.jpg",
    attribution = "© Government of Ontario",
    group = "Base layer: Aerial Imagery"
  ) %>%
  addRasterRGB(
    privateLandImage,
     r = 1, g = 2, b = 3,
    # stretch = "lin", maxBytes = 5e+07,
    group = "Base layer: Private Land"
  )%>%
# Areas
  addPolygons(
    data = studyArea,
    color = "red",
    weight = 3,
    opacity = 1,
    fillOpacity = 0,
    group = "Area: Study Area"    
  ) %>%
  addPolygons(
    data = surveyArea,
    color = c("red", "yellow"),
    weight = 1,
    opacity = 1,
    fillOpacity = 0.5,
    group = "Area:Survey Area",
    popup = c("Primary: 400km2", "Secondary: 350km2")
  ) %>%
  addPolygons(
    data = provincalPark,
    color = "green",
    weight = 1,
    opacity = 1,
    fillOpacity = 0.5,
    group = "Area:Provincal Park"
    # attribution = "© Government of Ontario"
  ) %>%
  addPolygons(
    data = township,
    color = "purple",
    weight = 2,
    opacity = 1,
    fillOpacity = 0.5,
    group = "Area:Township"
    # attribution = "© Government of Ontario"
  ) %>%
  addPolygons(
    data = lake,
    color = "blue",
    weight = 4,
    opacity =0,
    fillOpacity =0.8,
    popup = "Dog Lake",
    group = "Area:Lake"
    # attribution = "OpenStreetMap"
  ) %>%
# Infrastructure
  addPolygons(
    data = aerodrome,
    color = "yellow",
    weight = 1,
    opacity = 1,
    fillOpacity = 0.4,
    popup = "Aerodrome: Two Island Lake",
    group = "Aerodrome"
  ) %>%
  addCircles(
    data = buildings,
    radius = 1,
    opacity = 1,
    fillOpacity = 1,
    color = "brown",
    popup = "Building",
    group = "Infrastructure: Buildings"
    # attribution = "© Government of Ontario"
  ) %>%
  addCircles(
    data = recreationPoints,
    radius = 3,
    opacity = 1,
    fillOpacity = 1,
    color = "green",
    popup = "Recreation point",
    group = "Infrastructure: Recreation points"
    # attribution = "© Government of Ontario"
  ) %>%
  addCircles(
    data = towers,
    radius = 6,
    opacity = 1,
    fillOpacity = 1,
    color = "black",
    fillColor = "red",
    popup = "Tower",
    group = "Infrastructure: Towers"
    # attribution = "© Government of Ontario"
  ) %>%
  addPolylines(
    data = roads,
    color = "black",
    weight = 2,
    opacity = 0.8,
    group = "Infrastructure: Roads"
    # attribution = "OpenStreetMap"
  ) %>%
# Enviroment
  addPolygons(
    data = canopyCover["STRUCTUR_9"],
    fillColor = ~palGreen(canopyCover$STRUCTUR_9),
    weight = 1,
    opacity = 0,
    fillOpacity = 0.9,
    popup = ~paste0("Canopy Cover ", as.character(canopyCover$STRUCTUR_9),"%"),
    group = "Environment: SBFI 2020 Canopy Cover"
    # attribution = "Wulder et al. 2024"
  ) %>%
  addPolygons(
    data = mooseWinterAreas,
    color = "orange",
    weight = 1,
    opacity = 1,
    fillOpacity = 0.4,
    popup = "Moose wintering area",
    group = "Environment: Moose wintering area"
    # attribution = "© Government of Ontario"
  ) %>%
  addRasterImage(
    fracSnow,
    colors = palSnow,
    group = "Environment: Snow Cover",
    maxBytes = 50 * 1024 * 1024,
    attribution = "U.S. Geological Survey"
  ) %>%
  addRasterImage(
    LAI2021,
    colors = palLAI,
    group = "Environment: LAI Winter 2020-2021",
    maxBytes = 50 * 1024 * 1024,
    attribution = "Fernandes et al. 2023"
  ) %>%
    addRasterImage(
    LAI2122,
    colors = palLAI,
    group = "Environment: LAI Winter 2021-2022",
    maxBytes = 50 * 1024 * 1024,
    attribution = "Fernandes et al. 2023"
  ) %>%
    addRasterImage(
    LAI2223,
    colors = palLAI,
    group = "Environment: LAI Winter 2022-2023",
    maxBytes = 50 * 1024 * 1024,
    attribution = "Fernandes et al. 2023"
  ) %>%
    addRasterImage(
    LAI2324,
    colors = palLAI,
    group = "Environment: LAI Winter 2023-2024",
    maxBytes = 50 * 1024 * 1024,
    attribution = "Fernandes et al. 2023"
  ) %>%
    addRasterImage(
    LAI2425,
    colors = palLAI,
    group = "Environment: LAI Winter 2024-2025",
    maxBytes = 50 * 1024 * 1024,
    attribution = "Fernandes et al. 2023"
  ) %>%
   addRasterImage(
    forestLoss,
    colors = palLoss,
    group = "Environment: Tree Cover Loss 2001-2024",
    maxBytes = 50 * 1024 * 1024,
    attribution = "Hansen/UMD/Google/USGS/NASA"
  ) %>%
   addRasterImage(
    forestGain,
    colors = palGain,
    group = "Environment: Tree Cover Gain 2001-2024",
    maxBytes = 50 * 1024 * 1024,
    attribution = "Hansen/UMD/Google/USGS/NASA"
  ) %>%

  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions()
  ) %>%
  addMeasure(
  position = "bottomright",
  primaryLengthUnit = "kilometers",
  secondaryLengthUnit = "meters",
  primaryAreaUnit = "hectares", 
  secondaryAreaUnit = "sqmeters",
  activeColor = "#ABE67E",
  completedColor = "#C8F2BE",
  popupOptions = list(className = "leaflet-measure-resultpopup", autoPanPadding = c(10,
    10)),
  captureZIndex = 10000,
  localization = "en",
  decPoint = ".",
  thousandsSep = ","
) %>%
  addLegend("bottomleft", col = c("red", "green", "purple", "blue"), label = c("Survey area", "Provincal Park","Township", "Dog Lake"), title = "Area") %>%
  addLegend("bottomleft", col = c("yellow", "brown", "green", "red", "black"), label = c("Aerodrome: Two Island Lake", "Building", "Recreation point", "Tower", "Road"), title = "Infrastructure") %>%
  # addLegend("bottomleft", col = c(palGreen, palLAI, palSnow, "orange"), col = c(palGreen, palLAI, palSnow, "orange"), label = c("SBFI 2020 Canopy Cover", "LAI Winter", "Fraction Snow Cover", "Moose Wintering Area"), title = "Environment") %>%
  addLayersControl(
    overlayGroups = c("Base layer: OpenTopo", "Base layer: Mapzen Global Terrain", "Base layer: Relief map of DTM", "Base layer: Aerial Imagery", "Base layer: Private Land", "Area:Survey Area", "Area:Provincal Park", "Area:Township", "Area:Lake", "Aerodrome", "Infrastructure: Buildings", "Infrastructure: Recreation points", "Infrastructure: Towers", "Infrastructure: Roads", "Environment: SBFI 2020 Canopy Cover", "Environment: LAI Winter 2020-2021", "Environment: LAI Winter 2021-2022", "Environment: LAI Winter 2022-2023", "Environment: LAI Winter 2023-2024", "Environment: LAI Winter 2024-2025", "Environment: Tree Cover Gain 2001-2024", "Environment: Tree Cover Loss 2001-2024", "Environment: Snow Cover", "Environment: Moose wintering area"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  hideGroup(c("Base layer: OpenTopo", "Base layer: Mapzen Global Terrain", "Base layer: Relief map of DTM", "Base layer: Aerial Imagery", "Base layer: Private Land", "Area:Provincal Park", "Area:Township", "Area:Lake", "Aerodrome", "Infrastructure: Buildings", "Infrastructure: Recreation points", "Infrastructure: Towers", "Infrastructure: Roads", "Environment: SBFI 2020 Canopy Cover","Environment: Moose wintering area", "Environment: Snow Cover", "Environment: LAI Winter 2020-2021", "Environment: LAI Winter 2021-2022", "Environment: LAI Winter 2022-2023", "Environment: LAI Winter 2023-2024", "Environment: LAI Winter 2024-2025", "Environment: Tree Cover Gain 2001-2024", "Environment: Tree Cover Loss 2001-2024", "Environment: Snow Cover 2001-2024"))
```

## Private land map

[link](https://www.lioapplications.lrc.gov.on.ca/CLUPA/index.html?viewer=CLUPA.CLUPA&locale=en-CA){target="_blank"}




